Project Diary

### 2014-05-08 thesis/price_models.tex
Add the additions of stock case studies.

\section{Case Study: Volvo in 2013/14}
\label{sec:volvo}
In this section we model the log-volatility of {\it Volvo B} 30-minute
returns during the period 2013/10/10 - 2014/04/04. This series
contains 1884 log-volatilities computed using 2-minute returns in each
30-minute interval. Among them we use the first 1507 for model
estimation and the last 377 for forecast and model verification.

The left plot of figure \ref{fig:volvo_inno_and_lv_acf} shows the
distribution of $(r_t - \E(r_t))/\sigma_t$. We observe in the figure a
nice Gaussian variate, so we can be sure that the sum of squared
2-minute returns makes a good approximation to the variance of
30-minute returns in this particular case.
\begin{figure}[htb!]
  \centering
    \includegraphics[scale=0.5, clip=true, trim=0 256 0
    151]{../pics/volvo_inno_and_lv_acf.pdf}
    \label{fig:volvo_inno_and_lv_acf}
  \caption{\small \it Left: probability plot of $(r_t -
    \E(r_t))/\sigma_t$; Right: auto-correlations of $\ln \sigma_t$}
\end{figure}

Guided by the auto-correlations of $\ln\sigma_t$ shown in the right
plot of figure \ref{fig:volvo_inno_and_lv_acf} we find the following
model:
\begin{eqnarray}
  && (1 - B)(1 - B^s) \ln \sigma_t \nonumber \\
  &=& (1 - \theta_1 B - \theta_2B^2 -
  \theta_3B^3 - \theta_4B^4) (1 - \Theta B^s)
  y_t \label{eq:volvo_lv_model}
\end{eqnarray}
where $s = 16$ is the seasonality and is apparent from the
auto-correlations of $\ln\sigma_t$. Fitting this model to the
measured realized volatilities yields the parameter values listed in
table
\ref{tab:volvo_params}.
\begin{table}[htb!]
  \centering
  \begin{tabular}{|c|c|c|c|c|c|c|}
    \hline
    Parameter & $\theta_1$ & $\theta_2$ & $\theta_3$ & $\theta_4$ &
    $\Theta$ & residual variance \\
    \hline
    Value & 0.7305 & 0.0575 & 0.0574 & 0.0346 & 0.8324 & 0.1340\\
    \hline
  \end{tabular}
  \caption{\small \it Volvo B log-volatility parameters}
  \label{tab:volvo_params}
\end{table}
Forecasting using the estimated model parameters gives the forecast
series $\ln \sigma^{\text{SV}}_t$. To access the quality of the
forecast, we also estimate a GARCH model using the returns. The result
is a GARCH(1, 1) model, whose parameter values are listed in table
\ref{tab:volvo_garch}.
\begin{table}[htb!]
  \centering
  \begin{tabular}{|c|c|c|c|}
    \hline
    Parameter & $\alpha_0$ & $\alpha_1$  & $\beta_1$ \\
    \hline
    Value & $3.125 \times 10^{-7}$ & 0.05 & 0.90 \\
    \hline
  \end{tabular}
  \caption{\small \it GARCH(1, 1) model of Volvo B 30-minute returns}
  \label{tab:volvo_garch}
\end{table}

The forecasts from SV, GARCH, and the sample mean are compared
using the difference $\ln \sigma^F_t - \ln\hat{\sigma}_t$, where $\ln
\sigma^F_t$ stands for the forecast. The distributions of this
difference is plotted in figure \ref{fig:volvo_slv_garch_cmp}; the
mean and the standard deviation of the distributions are listed in
table \ref{tab:volvo_slv_garch_cmp}.
\begin{table}[htb!]
  \centering
  \begin{tabular}{|c|c|c|c|}
    \hline
    & SV & GARCH & Sample mean \\
    \hline
    $\E(\ln \sigma^F_t - \ln \hat{\sigma}_t)$ & -0.0123 &
    0.0242 & -0.1055 \\
    \hline
    $\text{std}(\ln \sigma^F_t - \ln \hat{\sigma}_t)$ & 0.3261 &
    0.4250 & 0.3708 \\
    \hline
  \end{tabular}
  \caption{\small \it Standard deviation of $\ln\sigma^F_t -
    \ln\hat{\sigma}_t$}
  \label{tab:volvo_slv_garch_cmp}
\end{table}

\begin{figure}[htb!]
  \centering
    \includegraphics[scale=0.5, clip=true, trim=10 282 0
    243]{../pics/volvo_slv_garch_cmp.pdf}
    \label{fig:volvo_slv_garch_cmp}
  \caption{\small \it Comparison of the forecast from SV, GARCH,
    and sample mean. Blue: SV forecast; Green: GARCH forecast; Red:
    Sample mean forecast.}
\end{figure}
Figure \ref{fig:volvo_slv_garch_cmp} shows, as in the previous case of
Nordea Bank 15-minute returns, the SV model performs the best, GARCH
the second, and the sample mean the worst. However, when it comes to
over-estimates, the sample mean appears to be the best estimator,
while SV excels over GARCH. But a check of $\E(\ln\sigma_t)$ over
the sample for estimation (1507 data points) and over the sample for
comparison (377 data points) reveals that the first sample has mean
-6.3127 while the second has -6.2072. This increment explains the low
probability of over-estimation when using the first sample mean as
forecast.

Table \ref{tab:volvo_good_percentage} compares the fraction of
``good'' estimates as measured by ${|\ln\sigma^F_t -
  \ln\hat{\sigma}_t| \over |\ln \hat{\sigma}_t|}$ being less than 1\%,
5\% and 10\%.
\begin{table}[htb!]
  \centering
  \begin{tabular}{|c|c|c|c|}
    \hline
    ${|\ln \sigma^F_t - \ln \hat{\sigma}_t| \over |\ln
      \hat{\sigma}_t|}$ &
    SV & GARCH & Sample Mean \\
    \hline
    1\% & 0.2175 & 0.1220 & 0.1406 \\
    \hline
    5\% & 0.7162 & 0.6154 & 0.6631 \\
    \hline
    10\% & 0.9231 & 0.8806 & 0.8992 \\
    \hline
  \end{tabular}
  \caption{\small \it Fraction of ``good'' forecasts as defined by
    ${|\ln \sigma^F_t - \ln \hat{\sigma}_t| \over |\ln
      \hat{\sigma}_t|}$ being less than 1\%, 5\% and 10\%.}
  \label{tab:volvo_good_percentage}
\end{table}
We see from the table that the SV model consistently excels over
the other two alternatives.

>>>>

\section{Case Study: Volvo in 2013/14}
\label{sec:volvo}
In this section we study the volatility of Volvo B 15-minute returns
during the period 2013/10/10 - 2014/04/04.
The subject of modeling is the realized volatility
computed as the square root of the sum of squared 1-minute
returns. The normal probability plot of the quotient $(r_t -
\E(r_t))/\sigma_t$ is shown in figure \ref{fig:volvo_15_quotient},
from which one can see it is normally distributed, confirming the
choice of 1-minute returns for computing the realized volatility.
\begin{figure}[htb!]
  \centering
  \includegraphics[scale=0.4, clip=true, trim=83 259 110 
  220]{../pics/volvo_15_quotient.pdf}
  \caption{\small \it Normal probability plot of $(r_t -
    \E(r_t))/\sigma_t$}
  \label{fig:volvo_15_quotient}
\end{figure}
The auto-correlations of $\ln \sigma_t$ and $w_t = (1-B)(1-B^s) \ln
\sigma_t$ are plotted in figure \ref{fig:volvo_15_lv_acf} and
\ref{fig:volvo_15_w_acf}. The former clearly shows the seasonality
$s = 33$ in the auto-correlations of $\ln \sigma_t$.

\begin{figure}[htb!]
  \centering
  \subfigure[]{
  \includegraphics[scale=0.4, clip=true, trim=90 259 103
  220]{../pics/volvo_15_lv_acf.pdf}
  \label{fig:volvo_15_lv_acf}
  }
  \subfigure[]{
  \includegraphics[scale=0.4, clip=true, trim=90 259 103
  220]{../pics/volvo_15_w_acf.pdf}
  \label{fig:volvo_15_w_acf}
  }
  \caption{\small \it Left: auto-correlations of $\ln \sigma_t$;
    Right: auto-correlations of $w_t = (1-B)(1-B^s) \ln \sigma_t$.}
\end{figure}

Based on the auto-correlations of $w_t$, a seasonal moving average
model is estimated:
\begin{eqnarray*}
  (1-B)(1-B^s) \ln\sigma_t &=& (1-\theta_1B - \theta_2B^2 - \theta_3
  B^3)(1 - \Theta B^s)y_t
\end{eqnarray*}
By maximum likelihood estimation, the parameter values listed in table
\ref{tab:volvo_15_sv_param} are obtained.
\begin{table}[htb!]
  \centering
  \begin{tabular}{|c|c|c|c|c|c|}
  \hline
  Parameter & $\theta_1$ & $\theta_2$ & $\theta_3$ & $\Theta$ & 
 $\var(y_t)$ \\
 \hline
 Value & 0.7491 & 0.1170 & 0.0571 & 0.8646 & 0.1245 \\
  \hline
  \end{tabular}
  \caption{\small \it Parameter values of the stochastoc volatility
    (SV) model.}
  \label{tab:volvo_15_sv_param}
\end{table}

A GARCH model (c.f. equation \ref{eq:garch_def}) assuming Gaussian
innovations is also fitted to the same series. The parameter values
are listed in table \ref{tab:volvo_15_garch_param}:
\begin{table}[htb!]
  \centering
  \begin{tabular}{|c|c|c|c|c|c|c|}
  \hline
  Parameter & $\alpha_0$ & $\alpha_1$ & $\beta_{1}$ \\
  \hline
  Value & $6.54 \times 10^{-7}$ & 0.1565 & 0.6445\\
  \hline
  \end{tabular}
  \caption{\small \it Parameter values of the GARCH model.}
  \label{tab:volvo_15_garch_param}
\end{table}

Performing one-step-ahead forecasts using both models gives the series
$\ln\sigma^F_t - \ln\hat{\sigma}_t$, where $\ln\hat{\sigma}_t$ are the
measured realized volatilities. The mean and standard deviation of
$\ln\sigma^F_t - \ln\hat{\sigma}_t$ are shown in table
\ref{tab:volvo_15_stat}.
\begin{table}[htb!]
  \centering
  \begin{tabular}{|c|c|c|c|}
    \hline
     & SV & GARCH & Sample mean \\
     \hline
    $\E(\ln \sigma^F_t - \ln \hat{\sigma}_t)$ & 0.0035 & 0.0313 & -0.1448 \\
    \hline
     $\text{std}(\ln \sigma^F_t - \ln \hat{\sigma}_t)$ & 0.3356 &
     0.3894 & 0.3826 \\
    \hline
 \end{tabular}
  \caption{\small \it Mean and standard deviation of the 3 kinds of
    forecasts}
  \label{tab:volvo_15_stat}
\end{table}
We see that both the SV and the GARCH model give biased forecasts, but
the SV forecasts are only 1/10 as biased as those of GARCH (0.0035
vs. 0.0313). In addition, the standard deviation of the SV forecasts
is smaller too. These results are confirmed by the distribution of
$\ln \sigma^F_t - \ln \hat{\sigma}_t$ and the fraction of ``good''
forecasts, which are shown in figure \ref{fig:volvo_15_diff} and table
\ref{tab:volvo_15_diff}, respectively.
\begin{figure}[htb!]
  \centering
    \includegraphics[scale=0.54, clip=true, trim=27 276 0
    236]{../pics/volvo_15_diff.pdf}
  \caption{\small \it Comparison of the forecasts from SV, GARCH,
    and sample mean. Blue: SV forecasts; Green: GARCH forecasts; Red:
    Sample mean forecasts.}
  \label{fig:volvo_15_diff}
\end{figure}
\begin{table}[htb!]
  \centering
  \begin{tabular}{|c|c|c|c|}
    \hline
    ${|\ln \sigma^F_t - \ln \hat{\sigma}_t| \over |\ln
      \hat{\sigma}_t|}$ &
    SV & GARCH & Sample Mean \\
    \hline
    1\% & 0.5760 & 0.5979 & 0.3930 \\
    \hline
    5\% & 0.8351 & 0.8363 & 0.6611 \\
    \hline
    10\% & 0.9716 & 0.9536 & 0.9085 \\
    \hline
  \end{tabular}
  \caption{\small \it Percentage of ``good'' forecasts as defined by
    deviating nore more than 1\%, 5\% and 10\% from the measured
    realized volatility.}
  \label{tab:volvo_15_diff}
\end{table}
